\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-25 src/tex/ms.tex
Adding comment on EOB fitting

Derivatives are ubiquitously useful throughout data analysis tasks.
% For instance, during PE, gradients are used to improve performance of the samplers 
% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
% \kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc. Check what method is used 
for IMRPhenomXPHM?}
\mi{I agree. it'd be good to comment on whether this is still done like this in
modern waveforms.}

@@

>>>>

Derivatives are ubiquitously useful throughout data analysis tasks.
% For instance, during PE, gradients are used to improve performance of the samplers 
% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo (HMC)~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically required a significant amount of work to obtain~\citep{Keppel:2013kia}.
With waveforms only increasing in complexity, calculating analytic derivatives will become ever more tricky.
% \kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

@@

%AZ: Addressed below
%\AZ{Maybe comment on optimization method used for EOB waveforms?
%I think they do interesting Monte Carlo etc. Check what method is used 
%for IMRPhenomXPHM?}
%\mi{I agree. it'd be good to comment on whether this is still done like this in
%modern waveforms.}

@@

\AZ{While we focused on \texttt{IMRPhenomD} here, the ability to apply AD to the calibration parameters may assist in other approaches to calibration, such as that used for the aligned-spin EOB model in~\citet{Bohe:2016gbl}.
There MCMC methods were used in a two-step procedure to optimize the calibration parameters. 
Derivative information, if implemented for EOB waveforms, could allow for the application of other sampling methods such as HMC, or possibly optimization over the entire set of NR waveforms at once.}
